---
title: "Summarise GDPDR opt-outs"
subtitle: "https://digital.nhs.uk/data-and-information/publications/statistical/national-data-opt-out/july-2021"
author: "Alasdair Henderson"
date: "30/07/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

library(pacman)
p_load(tidyverse)
p_load(ggplot2)
p_load(here)
p_load(gt)
#p_load(rgdal)
p_load(skimr)
p_load(maptools)
p_load(leaflet)
if (!require(gpclib)) install.packages("gpclib", type="source")
gpclibPermit()
theme_set(theme_minimal())

age_gen <- read_csv(here("data/NDOP_age_gen_Apr_2022.csv"))
reg <- read_csv(here("data/NDOP_reg_geo_Apr_2022.csv"))
res <- read_csv(here("data/NDOP_res_geo_Apr_2022.csv"))

#ccg_shape <- readOGR(here::here("data/Clinical_Commissioning_Groups_(April_2020)_EN_BFC_V2/Clinical_Commissioning_Groups_(April_2020)_EN_BFC_V2.shp"))

#regions_shape <- readOGR(here::here("data/NHS_England_Regions_(April_2019)_EN_BFC/NHS_England_Regions_(April_2019)_EN_BFC.shp"))

```

## Opt outs over time 

Note the big peak in opt outs across gender and age-bands in July 2021

```{r}
age_gen_edited <- age_gen %>%
  mutate_if(is.character, ~stringr::str_to_lower(.)) %>%
  janitor::clean_names() %>%
  filter(gender == "male" | gender == "female") %>%
  filter(!str_detect(age_band, "all")) %>%
  mutate_at("age_band", ~ifelse(. == "oct-19", "10-19", .)) %>%
  mutate(age_band = factor(age_band)) %>% 
  mutate_at("ach_date", ~lubridate::as_date(., format = "%d/%m/%Y"))

#age_gen_edited %>%
#  skim()

age_gen_edited %>%
  mutate(age_band_bigger = 
  fct_collapse(age_band, 
               "<20" = c("0-9", "10-19"),
               "20-40" = c("20-29", "30-39"),
               "40-60" = c("40-49", "50-59"),
               "60+" = c("60-69", "70-79", "80-89", "90+")
               )
  ) %>%
  group_by(ach_date, gender, age_band_bigger) %>% 
  summarise(opt_out = sum(opt_out), list_size = sum(list_size), .groups = "keep") %>% 
  mutate(rate = (opt_out/list_size)*100) %>%
  ggplot(aes(ach_date, rate, colour = age_band_bigger)) +
  geom_line() +
  labs(y = "drop out percent", x = "Date", colour = "Age group", title="Opt outs over time by gender and age group") +
  facet_wrap(~gender, ncol = 1) 

```

So the percentage of drop outs has changed over time, but the denominator (`list_size`) has not

```{r}
age_gen_edited <- age_gen %>%
  mutate_if(is.character, ~stringr::str_to_lower(.)) %>%
  janitor::clean_names() %>%
  filter(gender == "male" | gender == "female") %>%
  filter(!str_detect(age_band, "all")) %>%
  mutate_at("age_band", ~ifelse(. == "oct-19", "10-19", .)) %>%
  mutate(age_band = factor(age_band)) %>% 
  mutate_at("ach_date", ~lubridate::as_date(., format = "%d/%m/%Y"))

age_gen_edited %>%
  ggplot(aes(ach_date, list_size, colour = age_band)) +
  geom_line() +
  facet_wrap(~gender, ncol = 1)
```

Note the opt out count is the sum to date not the count in that month! 
```{r}
# age_gen_edited %>%
#   ggplot(aes(ach_date, opt_out, colour = age_band)) +
#   geom_line() +
#   facet_wrap(~gender, ncol = 1) 
```

So we just need to take the most recent data.

```{r}
opt_out_latest <- age_gen_edited %>% 
  filter(ach_date == max(ach_date))

```

### Gender

```{r}
## Gender
gender_summ <- opt_out_latest %>%
  group_by(gender) %>%
  summarise(opt_out = sum(opt_out), list_size = sum(list_size)) %>%
  mutate(pc = (opt_out/list_size)*100)

gt(gender_summ) %>%
  opt_align_table_header(align = "left") %>%
  fmt_number(columns = 2:3, decimals = 1, drop_trailing_zeros = T) %>%
  fmt_percent(columns = 4, decimals = 1, pattern = "({x})", scale_values = F)
```

Higher percentage of female opt outs (`r gender_summ$pc[gender_summ$gender == "female"] %>% round(digits = 1)`) than male (`r gender_summ$pc[gender_summ$gender == "male"] %>% round(digits = 1)`). 

### Age band
```{r}
## Age_band  
age_band_summ <- opt_out_latest %>%
  group_by(age_band) %>%
  summarise(opt_out = sum(opt_out), list_size = sum(list_size)) %>%
  mutate(pc = (opt_out/list_size)*100)

gt(age_band_summ) %>%
  opt_align_table_header(align = "left") %>%
  fmt_number(columns = 2:3, decimals = 1, drop_trailing_zeros = T) %>%
  fmt_percent(columns = 4, decimals = 1, pattern = "({x})", scale_values = F)
```

Highest in `r arrange(age_band_summ, -pc) %>% head(n=3) %>% pull(age_band)` groups.   
